Fail Faster

### 2025-03-19 paper/paper.tex
minor

\section{Evaluation}
\label{section:eval}

\jwc{Initial section here about eval}\tr{todo}

\begin{itemize}
  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
\end{itemize}

All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.

The OCaml version of \name and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
The Scala 3 version of \name and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.

@@

Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
\textit{strategies}, which vary in structure and sophistication. These include a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.

@@

These results show that staging and \rand choice serve distinct but complementary roles in generator performance. Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging. Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
\subsubsection{Scala}
To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala. We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').

As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups.

@@

It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance, or the ``speedup''. While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

>>>>

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. The OCaml version of \name and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
The Scala 3 version of \name and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.

To evaluate our contributions, this section answers the following research questions:
\begin{itemize}
  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
\end{itemize}

@@

Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
\textit{strategies}, which vary in structure and sophistication. These include type-derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.

@@

These results show that staging and \rand choice serve distinct but complementary roles in generator performance. Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging. Since both factors influence performance in different ways, generator libraries should use both in order to handle diverse workloads.
\subsubsection{Scala}
To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala. We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').

As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups in aggregate.

@@

It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-failure over a large number of trials. We normalize by computing the relative performance, or the ``speedup''. This works in most cases,
but it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}
